---
title: "GSEA Null Calibration: Permutation vs Rotation (Walkthrough)"
format:
  html:
    toc: true
    number-sections: true
---

[Portfolio](https://mirramockakid.github.io/Portfolio/){.chip-link}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 120,
  cache = TRUE
)

if (!requireNamespace("knitr", quietly = TRUE)) {
  stop("Package 'knitr' is required.")
}
```

# Overview

This document explains the main steps in `rotations.r`, runs the simulation script, and summarizes the differences between gene-wise permutation and rotation-based null generation.

The report is designed to rerun quickly:

- `rotations.r` now uses a simulation cache (`simulation_cache.rds`) for Monte Carlo outputs
- this Quarto document also enables knitr chunk caching

If the cache does not exist (or configuration changes), the first run can still take a while.

# What `rotations.r` Does (Step-by-Step)

## 1. Configuration

The script sets:

- dataset dimensions (`ngenes`, `nsamples`)
- gene set size (`set_size`, `set_idx`)
- Monte Carlo settings (`nsim`, `nperm`, `nrot`)
- correlation settings (`rho_set_grid`, `rho_bg`)
- parallelization mode (`parallel_over`, `ncores`)
- cache options (`use_simulation_cache`, `simulation_cache_file`)

## 2. Core helper functions

The script defines functions to:

- compute a GSEA-like enrichment score (`gsea_set_stat`)
- compute empirical right-tail p-values (`empirical_p_right`)
- simulate correlated noise in set and background genes
- simulate null datasets (no differential expression)
- fit limma and compute enrichment scores from p-values
- build a rotation basis from the null-model residual space
- sample random orthogonal rotations in that residual space

## 3. Null generation methods compared

For each simulated dataset, the script computes:

- **Permutation null**: shuffles gene-wise p-values before recomputing the set statistic
- **Rotation null**: rotates residuals under the null model, refits, and recomputes the set statistic

This is repeated over:

- multiple correlation strengths in `rho_set_grid`
- multiple Monte Carlo replicates (`nsim`)

## 4. Summaries and output files

The script then:

- summarizes empirical type-I error and mean p-values by `rho_set`
- writes CSV outputs
- creates diagnostic plots (calibration panel, ECDF, null histograms)
- creates correlation heatmaps and rotation-preservation diagnostics

# Run the Script (Using Cached Simulation Data)

```{r run-script, cache=TRUE}
if (!file.exists("rotations.r")) {
  stop("Cannot find rotations.r in the working directory.")
}

if (!file.exists("simulation_cache.rds")) {
  message("No simulation cache found yet. First run may be slow.")
}

sim_env <- new.env(parent = globalenv())
source("rotations.r", local = sim_env)

# Pull key objects into the report environment for convenience
needed <- c(
  "ngenes", "nsamples", "set_size", "set_idx", "nsim", "nperm", "nrot", "alpha",
  "rho_set_grid", "rho_bg", "parallel_over", "ncores",
  "results", "summary_tbl", "example_datasets", "example_distributions",
  "rotation_preservation_summary",
  "build_rotation_basis", "sample_rotation_matrix"
)
for (nm in needed) {
  if (exists(nm, envir = sim_env, inherits = FALSE)) {
    assign(nm, get(nm, envir = sim_env, inherits = FALSE))
  }
}
```

# Configuration Used in This Run

```{r config-table}
config_tbl <- data.frame(
  parameter = c(
    "ngenes", "nsamples", "set_size", "nsim", "nperm", "nrot", "alpha",
    "rho_set_grid", "rho_bg", "parallel_over", "ncores"
  ),
  value = c(
    ngenes, nsamples, set_size, nsim, nperm, nrot, alpha,
    paste(rho_set_grid, collapse = ", "),
    rho_bg, parallel_over, ncores
  ),
  stringsAsFactors = FALSE
)
knitr::kable(config_tbl, caption = "Configuration used by rotations.r")
```

# Simulation Outputs

## Monte Carlo results (head)

```{r results-head}
knitr::kable(head(results, 10), caption = "First 10 Monte Carlo rows")
```

## Calibration summary by correlation level

```{r summary-table}
summary_view <- summary_tbl
summary_view$delta_type1_perm_minus_rot <- summary_view$type1_perm - summary_view$type1_rot
summary_view$delta_mean_p_perm_minus_rot <- summary_view$mean_p_perm - summary_view$mean_p_rot
knitr::kable(summary_view, digits = 4, caption = "Calibration summary across rho_set")
```

# Visualizing the Difference: Permutation vs Rotation

## 1. Type-I error calibration vs within-set correlation

This reproduces the main "Calibration Under Null" comparison and highlights how the methods behave as gene-gene correlation increases.

```{r calibration-plot}
plot_tbl <- summary_tbl[
  is.finite(summary_tbl$rho_set) &
    is.finite(summary_tbl$type1_perm) &
    is.finite(summary_tbl$type1_rot),
, drop = FALSE]
plot_tbl <- plot_tbl[order(plot_tbl$rho_set), , drop = FALSE]

if (nrow(plot_tbl) == 0) {
  plot.new()
  title("Calibration Under Null")
  text(0.5, 0.5, "No finite values to plot")
} else {
  y_lim <- range(c(plot_tbl$type1_perm, plot_tbl$type1_rot, alpha), na.rm = TRUE)
  if (y_lim[1] == y_lim[2]) y_lim <- y_lim + c(-0.01, 0.01)

  plot(
    plot_tbl$rho_set, plot_tbl$type1_perm,
    type = "b", pch = 16, lwd = 2,
    ylim = y_lim,
    xlab = expression(rho[set]),
    ylab = "Empirical type-I error",
    main = "Calibration Under Null"
  )
  lines(plot_tbl$rho_set, plot_tbl$type1_rot, type = "b", pch = 17, lwd = 2, col = "blue")
  abline(h = alpha, lty = 2, col = "darkgray")
  legend(
    "topleft",
    legend = c("Gene-wise permutation", "Rotation", "Nominal alpha"),
    col = c("black", "blue", "darkgray"),
    lty = c(1, 1, 2),
    pch = c(16, 17, NA),
    bty = "n"
  )
}
```

## 2. p-value ECDF comparison at low and high correlation

Uniformity under the null is a useful diagnostic. The dashed line is the ideal uniform ECDF.

```{r ecdf-comparison, fig.width=10, fig.height=4.5}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

for (rho_val in c(min(rho_set_grid), max(rho_set_grid))) {
  idx <- results$rho_set == rho_val & is.finite(results$p_perm) & is.finite(results$p_rot)
  if (!any(idx)) {
    plot.new()
    title(sprintf("ECDF (rho_set = %.1f)", rho_val))
    text(0.5, 0.5, "No finite p-values")
  } else {
    plot(
      ecdf(results$p_perm[idx]),
      do.points = FALSE,
      verticals = TRUE,
      main = sprintf("ECDF (rho_set = %.1f)", rho_val),
      xlab = "p-value",
      ylab = "ECDF"
    )
    lines(ecdf(results$p_rot[idx]), do.points = FALSE, verticals = TRUE, col = "blue")
    abline(0, 1, lty = 2, col = "darkgray")
    legend(
      "topleft",
      legend = c("Permutation", "Rotation", "Uniform(0,1)"),
      col = c("black", "blue", "darkgray"),
      lty = c(1, 1, 2),
      bty = "n"
    )
  }
}
par(mfrow = c(1, 1))
```

## 3. Example null distributions of the enrichment statistic

These panels compare the null distributions produced by the two methods for a single example dataset at low and high correlation.

```{r null-distribution-panels, fig.width=10, fig.height=4.5}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot_example_null <- function(ex, rho_label) {
  if (is.null(ex)) {
    plot.new()
    title(sprintf("Example null ES (rho_set = %s)", rho_label))
    text(0.5, 0.5, "Not available")
    return(invisible(NULL))
  }
  hist(
    ex$null_perm, breaks = 40, col = "gray85", border = "white",
    main = sprintf("Example null ES (rho_set = %s)", rho_label),
    xlab = "ES"
  )
  hist(ex$null_rot, breaks = 40, col = rgb(0, 0, 1, 0.35), border = "white", add = TRUE)
  abline(v = ex$ES_obs, col = "red", lwd = 2)
  legend(
    "topright",
    legend = c("Permutation", "Rotation", "Observed ES"),
    fill = c("gray85", rgb(0, 0, 1, 0.35), NA),
    border = NA,
    lty = c(NA, NA, 1),
    col = c(NA, NA, "red"),
    bty = "n"
  )
}

ex_low <- example_distributions[[paste0("rho_", min(rho_set_grid))]]
ex_high <- example_distributions[[paste0("rho_", max(rho_set_grid))]]

plot_example_null(ex_low, sprintf("%.1f", min(rho_set_grid)))
plot_example_null(ex_high, sprintf("%.1f", max(rho_set_grid)))

par(mfrow = c(1, 1))
```

# Why Rotation Helps: Correlation Structure Preservation

The key idea is that rotation operates in the residual space and preserves the covariance/correlation structure more faithfully than gene-wise permutation.

## 4. Correlation heatmaps from the simulated datasets

These are the summary statistics generated by `rotations.r` from example datasets.

```{r heatmap-summary}
if (file.exists("corr_heatmap_summary.csv")) {
  corr_heatmap_summary <- read.csv("corr_heatmap_summary.csv")
  knitr::kable(corr_heatmap_summary, digits = 4, caption = "Correlation heatmap summary statistics")
} else {
  cat("`corr_heatmap_summary.csv` not found. Re-run `rotations.r`.\n")
}
```

## 5. Rotation correlation-preservation metrics

This directly quantifies how similar the pairwise gene-gene correlations are before vs after a residual-space rotation.

```{r rotation-preservation-summary}
if (exists("rotation_preservation_summary")) {
  knitr::kable(rotation_preservation_summary, digits = 4, caption = "Rotation correlation-preservation summary")
} else if (file.exists("rotation_corr_preservation_summary.csv")) {
  rotation_preservation_summary <- read.csv("rotation_corr_preservation_summary.csv")
  knitr::kable(rotation_preservation_summary, digits = 4, caption = "Rotation correlation-preservation summary")
} else {
  cat("Rotation preservation summary not available.\n")
}
```

## 6. Inline correlation-preservation visualization (original vs rotated)

The next chunk recreates the key idea visually for the lowest and highest `rho_set` example datasets:

- original gene-gene correlation heatmap
- rotated gene-gene correlation heatmap
- scatter of pairwise correlations (original vs rotated)

```{r rotation-preservation-inline, fig.width=11, fig.height=8}
plot_rotation_preservation_inline <- function(Y, set_idx, rho_val, bg_n = 300) {
  if (is.null(Y)) {
    plot.new()
    title(sprintf("rho_set = %.1f", rho_val))
    text(0.5, 0.5, "Example dataset not available")
    return(invisible(NULL))
  }

  nsamp <- ncol(Y)
  design0 <- matrix(1, nrow = nsamp, ncol = 1)
  fit0 <- limma::lmFit(Y, design0)
  Yhat0 <- fitted(fit0)
  res0 <- Y - Yhat0
  basis <- build_rotation_basis(design0 = design0, nsamples = nsamp)
  Rmat <- sample_rotation_matrix(basis)
  Y_rot <- Yhat0 + res0 %*% Rmat

  bg_idx <- setdiff(seq_len(nrow(Y)), set_idx)
  if (length(bg_idx) > bg_n) bg_idx <- sample(bg_idx, bg_n)
  keep_idx <- c(set_idx, bg_idx)

  corr_orig <- cor(t(Y[keep_idx, , drop = FALSE]))
  corr_rot <- cor(t(Y_rot[keep_idx, , drop = FALSE]))
  u <- upper.tri(corr_orig)
  x <- corr_orig[u]
  y <- corr_rot[u]
  r_xy <- suppressWarnings(cor(x, y, use = "complete.obs"))
  mad_xy <- mean(abs(x - y), na.rm = TRUE)

  cols <- colorRampPalette(c("#313695", "#ffffbf", "#a50026"))(100)

  image(
    seq_len(nrow(corr_orig)), seq_len(ncol(corr_orig)),
    corr_orig[nrow(corr_orig):1, ],
    col = cols, zlim = c(-1, 1), axes = FALSE,
    xlab = "Genes", ylab = "Genes",
    main = sprintf("Original (rho=%.1f)", rho_val)
  )
  box()

  image(
    seq_len(nrow(corr_rot)), seq_len(ncol(corr_rot)),
    corr_rot[nrow(corr_rot):1, ],
    col = cols, zlim = c(-1, 1), axes = FALSE,
    xlab = "Genes", ylab = "Genes",
    main = "Rotated"
  )
  box()

  plot(
    x, y,
    pch = 16, cex = 0.35,
    col = rgb(0, 0, 0, 0.18),
    xlab = "Original pairwise correlation",
    ylab = "Rotated pairwise correlation",
    main = "Pairwise correlation comparison"
  )
  abline(0, 1, col = "red", lwd = 2)
  mtext(sprintf("r = %.3f | mean abs diff = %.3f", r_xy, mad_xy), side = 3, line = 0.2, cex = 0.8)
}

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
plot_rotation_preservation_inline(example_datasets[[paste0("rho_", min(rho_set_grid))]], set_idx, min(rho_set_grid))
plot_rotation_preservation_inline(example_datasets[[paste0("rho_", max(rho_set_grid))]], set_idx, max(rho_set_grid))
par(mfrow = c(1, 1))
```

# Interpretation Guide

When you read the report, the main differences to focus on are:

- **Type-I error calibration**:
  - if permutation inflates type-I error as `rho_set` increases, it suggests correlation is not being respected by the null
  - rotation should remain closer to nominal `alpha`
- **p-value ECDFs**:
  - departures from the diagonal under the null indicate miscalibration
- **Null ES histograms**:
  - compare how the null distributions differ in spread/shape
- **Correlation-preservation plots**:
  - points near the `y = x` line indicate rotated data preserve pairwise correlations well

# Files Produced by `rotations.r`

```{r produced-files}
produced_files <- c(
  "simulation_cache.rds",
  "null_calibration_summary.csv",
  "null_calibration_results.csv",
  "null_calibration.pdf",
  "corr_heatmap_summary.csv",
  "rotation_corr_preservation_summary.csv"
)

file_status <- data.frame(
  file = produced_files,
  exists = file.exists(produced_files),
  stringsAsFactors = FALSE
)
knitr::kable(file_status, caption = "Key output files")

cat("\nHeatmap PDFs:", length(Sys.glob("corr_heatmap_rho_*.pdf")), "\n")
cat("Rotation-preservation PDFs:", length(Sys.glob("rotation_corr_preservation_rho_*.pdf")), "\n")
```
